A causal analysis of the effect of abortion policies on maternal mortality

### 2022-06-08 PaperAnalysis.Rmd
Final edits

@@

Abortion law and the negative consequences of the lack of safe abortions are highly contentious policy issues that have been making headlines this year. Further, the [Sustainable Development Goal indicator 3.1](https://www.who.int/data/gho/indicator-metadata-registry/imr-details/26) of the United Nations to reduce Maternal Death to less than 70 per 100000 life birth by 2030 relies on access to safe and legal abortions. In the paper we are analysing, the authors are testing whether and, if so, to what extent abortion law as a proxy for access to safe abortions causally affects maternal mortality rates in over 160 countries around the globe. Establishing a causal relation could strengthen the implementation of abortion liberalisation policy. This is especially needed in a time where several states already withdrew or are withdrawing access to safe abortions on many grounds, such as [Poland](https://www.amnesty.org/en/latest/news/2021/10/poland-a-year-on-abortion-ruling-harms-women/) and the [USA](https://www.euronews.com/2022/05/03/us-supreme-court-moving-to-overturn-key-abortion-rights-law-says-leaked-report). While on the other hand there are developments like, for instance, in [France](https://www.hrw.org/news/2022/03/01/france-expands-abortion-access-two-key-moves) to expand abortion rights in 2022. So does abortion law flexibility cause maternal mortality? We will be taking a close critical look at a paper that tries to establish exactly that and also run our own data analysis to address this question.

@@

In "Abortion laws reform may reduce maternal mortality: an ecological study in 162 countries," @Latt2019 present the results of a study that aims to assess whether maternal mortality is affected by the flexibility of abortion laws. Abortion law flexibility was measured as a sum of the number of reasons for which abortion is legally allowed (ranging from 0 to 7) and was included in models as both an ordinal variable and a binary variable (inflexible vs. flexible abortion laws). Abortion laws were considered inflexible if there were less than 3 legal grounds for abortion. Maternal mortality ratio (MMR) was measured as the number of maternal deaths per 100,000 live births. Authors used fixed effects linear regression models to estimate the relationship between abortion law flexibility and MMR, while controlling for GDP per capita and time in five-year intervals. Additional analysis included also female primary education completion as a control variable as it had over 50% of missing data; however, results did not change. Results indicated that maternal deaths reduce by 45 per 100,000 live births (95% CI: -64, -26) with flexible abortion laws. The authors conclude that abortion law reform in countries with inflexible abortion laws may reduce maternal mortality.

@@

The authors state that they aim to “estimate the association between the flexibility of abortion laws and maternal mortality at a country level [between 1985 and 2013]” [@Latt2019, pg. 2]. Furthermore, they estimated the model using linear regression, and the quantity they have estimated was the regression coefficient of MMR on flexibility of the law. Therefore, we believe that the authors wanted to assess the average treatment effect (ATE), that is, the difference in the number of maternal deaths that would occur within a country if abortion law was flexible ($MMR_i$ = 1) versus inflexible ($MMR_i$ = 0):

@@

This estimand looks at counterfactual outcomes for all countries, regardless of the fact that the particular country would in reality receive treatment or not. Country $i$'s MMR when abortion laws are flexible is denoted as $MMR_i^{(L=1)}$ or shortly $MMR_i^{(1)}$, and as $MMR_i^{(L=0)}$ or $MMR_i^{(0)}$ when abortion laws are inflexible. Individual causal effect for the country $i$ is then the difference between mortality when abortion laws are flexible vs. when they are inflexible, or $ΔMMR_i = MMR_i^{(1)} - MMR_i^{(0)}$. Individual effects averaged across all of the countries in the world would then give the wanted causal quantity. This is equal to the expectation of the effect of flexible abortion laws on MMR of a randomly selected country.

@@

However, there are several potential identification issues that could influence the possibility to estimate the effect of abortion law flexibility on MMR, namely: unmeasured confounding and (non-random) data missingness. These are discussed in detail below.

@@

@@

(Non-random) Data missingness is also a factor that can affect the possibility of accurately identifying causal effects.In the original paper, Hernan et al. are outlining why data is missing. Because restricting the analysis on countries where data are complete would bias it, the authors still include countries with missing data on female literacy [@HernanSelection, p. 5]. However, female primary completion is only available for 50 percent of countries and female literacy only for 10 percent. This missingness of data could introduce selection bias to the specific analysis the auhtors did in the original paper.

Comparing the rate of MMR and flexibility score for countries with recorded female primary completion vs countries without recorded female primary completion levels, we could check whether there is a selection bias and then generalization of effects of this model would not be possible. If we suppose that female literacy is a collider and a confounder, we cannot condition on it without creating selection bias [@elwert2014endogenous, p.38]. As this done in the baseline paper, we would have to select strategies to correct this. Otherwise, it would stay d-connected, so statistically dependent.

@@

![Figure 4.:Female literacy's effect on Maternal mortality ratio](Figures/Dag_model3.png)

@@

![Figure 5.: Suggested causal model for future research ](Figures/Dag_model4.png)

* Sensitivity Analysis

In the sensitivity analysis we would check which part of the total effect can be explained by the abortion laws and which can be explained by female literacy. Additionally, we check how much the effects would need to change to cancel one another.For instance, the effect for flexibility score alone we can check when we control (block) the primary completion rate.
Sensitivity analysis has 2 essential parts to it:
1) describing the type of unobserved confounders
Unmeasured variables: religious institutions, contraceptive prevalence, fertility rate 
2) plausibility that problematic confounding might exist
	Based on 
	*research design, 
  *expert knowledge and 
  *data-generating process

Generally, sensitivity analysis is still underutilized in social science and policy research papers [@cinelli2020making, p. 40]. However, it is a very useful tool to test robustness of findings, i.e. how sensitive my effects and effect sizes are to other variables that influence the treatment or outcome studied. Therefore, we are proposing to add a sensitivity analysis to the research design and also apply it in the second part of our report. This method gives us two insights: 1) the robustness value: how robust the coefficient is to confounders influencing treatment and outcome and 2) the proportion of the outcome uniquely explained by the treatment and how strong a confounder would have to be to eliminate the effect of the treatment[@cinelli2020making, pp.40/41].

@@

* Heterogeneous effects

@Aronov2016 showed that multiple regression estimates are actually weighted averages of unit-specific contributions. In cases when treatment effects are heterogeneous, in the sense that the effect of abortion law flexibility is not the same for every country, $\beta_L$ will not reflect ATE. Rather, it will be more strongly influenced by units that have more variability: many observations, large variations in abortion law flexibility and MMR, and that are hard to predict from country and time fixed effects and other controls (GDP, female primary education completion). This means that the estimated effect will be actually valid for a *sub-sample* of the population that is being studied. @Aronov2016 suggest that the *effective* sample, the sub-sample that has the strongest weights in the analysis should be reported explicitly so that it is more clear to which cases the effects generalise.

* Missingness of counterfactuals.

@@

*[Add table showing median per period.]*

@@

1. regressing the confounders and the fixed effects for countries in the analysis (without the treatment) and the time period on the abortion law flexibility score (treatment measured as a score from 0 to 7) 
2. squaring the weights.

As expected, our analysis showed that countries have different weights in the calculated effect. The nominal sample, the full population of the countries that have data for both the treatment and the outcome, is shown on Figure 6a and it includes almost all countries in the world. However, the effective sample, defined here as the countries that contribute over half (51%) of the weight used to construct the estimate of the abortion law flexibility on MMR, consists of only 14 countries (Figure 6b).

@@

Finally, we conducted a sensitivity analysis to estimate the fragility of our the effect of interest against unobserved counfounders in our model[@cinelli2020making]. These potential confounders are not usually considered in  traditional regression estimates to avoid bias, but they might substantively change our conclusions.

@@

```{r, echo=FALSE,out.width="49%",out.height="49%",fig.show='hold', fig.align='center', fig.cap="Figure X. a) Countour plot for the sensitivity analysis of model 2 using primary completion level as benchmark confounder vs. b) Countour plot for the sensitivity analysis of model 1 using gdp as benchmark confounder"}

@@

## Conclusion 
The policy conclusion the original paper draws is justified. However, we could expand upon the methodology to get a more precise result and consider possible models as the limitation of the approach is not sufficiently justified in their paper. There are several ways to amend or extend the original research design. We presented suggestions in the theoretical part of our report like applying sensitivity analysis and accounting for the heterogeneity of effects. We generated partial insights with our own data analysis. The causal relation between abortion law and maternal mortality ratios could be established. However, the proportion of the effect remains to be studied with an extended methodological toolbox.

@@

We worked together as a team where we all worked on all parts and engaged with our colleagues' ideas through out. Each of us took the lead on parts of the report as detailed below:

* Helena Bakic : Empirical estimation,Identification, Heterogeinity of effects, Analysis
* Anna-Lisa Wirth (213286@mds.hertie-schoo.org) : Background, Unmeasured Confounding, Missing Data, Sensitivity Analysis, Analysis and Conclusion 
* Ángela Duarte Pardo : Data Cleaning, G-separation, Methodology, Modelling and Analysis

>>>>

Word count: app. 4637 words

@@

Abortion law and the negative consequences of the lack of safe abortions are highly contentious policy issues that have been making headlines this year. Further, the [Sustainable Development Goal indicator 3.1](https://www.who.int/data/gho/indicator-metadata-registry/imr-details/26) of the United Nations to reduce Maternal Death to less than 70 per 100,000 live births by 2030 relies on access to safe and legal abortions. In the paper we are analysing, the authors are testing whether and, if so, to what extent abortion law as a proxy for access to safe abortions causally affects maternal mortality rates in over 160 countries around the globe. Establishing a causal relation could strengthen the implementation of abortion liberalisation policy. This is especially needed in a time where several states already withdrew or are withdrawing access to safe abortions on many grounds, such as [Poland](https://www.amnesty.org/en/latest/news/2021/10/poland-a-year-on-abortion-ruling-harms-women/) and the [USA](https://www.euronews.com/2022/05/03/us-supreme-court-moving-to-overturn-key-abortion-rights-law-says-leaked-report). Conversly, some other countries, for instance, [France](https://www.hrw.org/news/2022/03/01/france-expands-abortion-access-two-key-moves) are planning to expand abortion rights in 2022. So does abortion law flexibility cause maternal mortality? We will be taking a close critical look at a paper that tries to establish exactly that and also run our own data analysis to address this question.

@@

In "Abortion laws reform may reduce maternal mortality: an ecological study in 162 countries," @Latt2019 present the results of a study that aims to assess whether maternal mortality is affected by the flexibility of abortion laws. Abortion law flexibility was measured as a sum of the number of reasons for which abortion is legally allowed (ranging from 0 to 7) and was included in models as both an ordinal variable and a binary variable (inflexible vs. flexible abortion laws). Abortion laws were considered inflexible if there were less than 3 legal grounds for abortion. Maternal mortality ratio (MMR) was measured as the number of maternal deaths per 100,000 live births. Authors used fixed effects linear regression models to estimate the relationship between abortion law flexibility and MMR, while controlling for GDP per capita, country variability and time. Additional analysis included female primary education completion as a potential control variable that was not included in the main model as it had over 50% of missing data; however, results did not significantly change. Results indicated that maternal deaths reduce by 45 per 100,000 live births (95% CI: -64, -26) with flexible abortion laws. The authors conclude that abortion law reform in countries with inflexible abortion laws may reduce maternal mortality.

@@

The authors state that they aim to “estimate the association between the flexibility of abortion laws and maternal mortality at a country level [between 1985 and 2013]” [@Latt2019, pg. 2]. Furthermore, they estimated the model using linear regression, and the quantity they have estimated was the regression coefficient of MMR on the abortion law flexibility. Therefore, we believe that the authors wanted to assess the average treatment effect (ATE), that is, the difference in the number of maternal deaths that would occur within a country if abortion law was flexible versus inflexible:

@@

This estimand looks at counterfactual outcomes for all countries, regardless of whether or not a particular country would in reality receive the treatment. Country $i$'s MMR when abortion laws are flexible is denoted as $MMR_i^{(L=1)}$ or shortly $MMR_i^{(1)}$, and as $MMR_i^{(L=0)}$ or $MMR_i^{(0)}$ when abortion laws are inflexible. Individual causal effect for the country $i$ is then the difference between mortality when abortion laws are flexible vs. when they are inflexible, or $ΔMMR_i = MMR_i^{(1)} - MMR_i^{(0)}$. Individual effects averaged across all of the countries in the world would then give the wanted causal quantity. This is equal to the expectation of the effect of flexible abortion laws on MMR of a randomly selected country.

@@

However, there are several potential identification issues that could influence the possibility to estimate the effect of abortion law flexibility on MMR, namely: unmeasured confounding and (non-random) data missingness. These are discussed in more detail below.

@@

To check the robustness of the findings authors could have examined how big the unobserved confounding would have to be to cancel the estimated causal effect. Sensitivity analysis has 2 essential parts to it: 1) describing the type of unobserved confounders such as religious institutions, contraceptive prevalence, fertility rate, 2) describing the plausibility that problematic confounding might exist based on research design, expert knowledge and data-generating process.

Generally, sensitivity analysis is still underutilized in social science and policy research papers [@cinelli2020making, p. 40]. However, it is a very useful tool to test robustness of findings, i.e. how sensitive my effects and effect sizes are to other variables that influence the treatment or outcome studied. Therefore, we are proposing to add a sensitivity analysis to the research design and also apply it in the second part of our report. This method gives us two insights: 1) the robustness value: how robust the coefficient is to confounders influencing treatment and outcome and 2) the proportion of the outcome uniquely explained by the treatment and how strong a confounder would have to be to eliminate the effect of the treatment [@cinelli2020making, pp.40/41].

@@

(Non-random) Data missingness is also a factor that can affect the possibility of accurately identifying causal effects [@HernanSelection]. Because restricting the analysis on countries where data are complete would bias it, the authors still include countries with missing data on female literacy, another possible confounder [@Latt2019, p. 5]. However, female primary completion is only available for 50 percent of countries and female literacy only for 10 percent. This missingness of data could introduce selection bias to the specific analysis the authors did in the original paper.

Comparing the rate of MMR and flexibility score for countries with recorded female primary completion vs countries without recorded female primary completion levels, we could check whether there is a selection bias and then generalization of effects of this model would not be possible. If we suppose that female literacy is a collider and a confounder, we cannot condition on it without creating selection bias [@elwert2014endogenous, p.38]. As this done in the current paper, we would have to select strategies to correct this. Otherwise, it would stay d-connected, so statistically dependent.

@@

```{r, echo=FALSE,out.width="49%",out.height="49%", fig.align='center', fig.cap="Figure 4.:Female literacy's effect on Maternal mortality ratio"}

knitr::include_graphics("Figures/Dag_model3.png")
```

@@

```{r, echo=FALSE,out.width="60%",out.height="60%", fig.align='center', fig.cap="Figure 5.: Suggested causal model for future research"}

knitr::include_graphics("Figures/Dag_model4.png")
```

@@

@Aronov2016 showed that multiple regression estimates are actually weighted averages of unit-specific contributions. In cases when treatment effects are heterogeneous, for example if the effect of abortion law flexibility is not the same for every country in the current paper, $\beta_L$ will not reflect ATE. Rather, it will be more strongly influenced by units that have more variability: many observations, large variations in abortion law flexibility and MMR, and that are hard to predict from country and time fixed effects and other controls (GDP, female primary education completion). This means that the estimated effect will be actually valid for a *sub-sample* of the population that is being studied. @Aronov2016 suggest that the *effective* sample, the sub-sample that has the strongest weights in the analysis should be reported explicitly so that it is more clear to which cases the effects generalise.

@@

@@

1. regressing the GDP and the fixed effects for countries and time period on the abortion law flexibility score (measured as a score from 0 to 7); 
2. squaring the weights.

As expected, our analysis showed that countries have different weights in the estimated causal effect. The nominal sample, the full population of the countries that have data for both the treatment and the outcome, is shown on Figure 6a and it includes almost all countries in the world. However, the effective sample, defined here as the countries that contribute over half (51%) of the weight used to construct the estimate of the abortion law flexibility on MMR, consists of only 14 countries (Figure 6b).

@@

Finally, we conducted a sensitivity analysis to estimate the robustness of the effect of interest against unobserved counfounders in our model [@cinelli2020making]. These potential confounders are not usually considered in  traditional regression estimates to avoid bias, but they might substantively change our conclusions.

@@

```{r, echo=FALSE,out.width="49%",out.height="49%",fig.show='hold', fig.align='center', fig.cap="Figure 8. a) Countour plot for the sensitivity analysis of model 2 using primary completion level as benchmark confounder vs. b) Countour plot for the sensitivity analysis of model 1 using gdp as benchmark confounder"}

@@

## Conclusion 
The policy conclusion the original paper draws is justified. However, we could expand upon the methodology to get a more precise result and consider possible models as the limitation of the approach is not sufficiently justified in their paper. There are several ways to amend or extend the original research design. We presented suggestions in the theoretical part of our report like applying sensitivity analysis and accounting for the heterogeneity of effects. We generated partial insights with our own data analysis. The causal relation between abortion law and maternal mortality ratios could be established. However, the strength of the effect remains to be studied with an extended methodological toolbox.

@@

We worked together as a team on all parts of the paper and engaged with our colleagues' ideas throughout. Each of us took the lead on parts of the report as detailed below:

* Helena Bakic (bakic.helena@outlook.com): Empirical estimation, Identification, Heterogeinity of effects, Analysis
* Anna-Lisa Wirth (213286@mds.hertie-schoo.org): Background, Unmeasured Confounding, Missing Data, Sensitivity Analysis, Analysis and Conclusion 
* Ángela Duarte Pardo: Data Cleaning, G-separation, Methodology, Modelling and Analysis
